---
title: "Animal trajectory analysis with trajr"
author: "Jim McLean"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to trajr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library("plotrix")
library(trajr)
```

Trajr is an R toolkit for the statistical analysis of animal trajectories.

## Trajectories

Trajr is used to analyse 2-dimensional trajectories with an additional time component. A trajectory can also be thought of as a series of steps, each with a length and a turning angle.

```{r fig.cap="A simple trajectory", fig.height=8, fig.width=6}
par(mar=c(4, 4, 3, 1), mfrow = c(2, 1))
n <- 5

# Generate a random walk trajectory
set.seed(4)
trj <- TrajGenerate(n, random = TRUE)
plot(trj, main = "Random walk", turning.angles = "random")

# Generate a random directed walk trajectory
set.seed(1)
trj <- TrajGenerate(n, random = FALSE)
plot(trj, main = "Directed walk", turning.angles = "directed")

```

## Expected results of navigation

Plot the expected results of navigation with and without a compass, based on (Cheung, Zhang, Stricker, & Srinivasan, 2007).

```{r directWalks, fig.cap="Simulated directed journeys", fig.height=5, fig.width=6, cache=TRUE}
simulateJourneys <- function(random, nTrajectories = 10000) {
  trjs <- lapply(1:nTrajectories, function(i) TrajGenerate(n = 500, random = random))
  par(mfrow = c(2, 3), mar = c(4, 4, 3, 1))
  for(step in c(1, 2, 5, 20, 100, 500)) {
    title <- sprintf("%d %s", step, ifelse(step == 1, "step", "steps"))
    plot(t(sapply(trjs, function(t) t[step + 1,c('x', 'y')])), pch  = '.', main = title, asp = 1)
  }
}
simulateJourneys(random = FALSE)
```

```{r randomWalks, fig.cap="Simulated random journeys", fig.height=5, fig.width=6, cache=TRUE}
simulateJourneys(random = TRUE)
```


## E[max]

Some empirical testing of Emax.

```{r Emax, fig.cap="E[max] as a function of angular error", cache=TRUE, fig.height=4, fig.width=6}
set.seed(1)
n <- 5000
angularErrorSd <- runif(n, 0, 2)

# Assess Emax for some different types of trajectories, all with the same angular error sd
emaxr <- sapply(1:n, function(i) { TrajEmax(TrajGenerate(500, random = TRUE, angularErrorSd = angularErrorSd[i])) })
emaxd <- sapply(1:n, function(i) { TrajEmax(TrajGenerate(500, random = FALSE, angularErrorSd = angularErrorSd[i]), compass.direction = 0) })
emaxSm <- sapply(1:n, function(i) { TrajEmax(TrajRediscretize(TrajGenerate(500, random = TRUE, angularErrorSd = angularErrorSd[i]), R = 1)) })
emaxBig <- sapply(1:n, function(i) { TrajEmax(TrajRediscretize(TrajGenerate(500, random = TRUE, angularErrorSd = angularErrorSd[i]), R = 10)) })
emaxBig[emaxBig < 0] <- NA # can't take log of negative numbers

par(mar = c(4, 4, 1, 1))
plot(rep(angularErrorSd, 4), c(emaxr, emaxd, emaxSm, emaxBig),
     log = 'xy', pch = '.', 
     col = c(rep('red', n), rep('blue', n), rep('green', n), rep('black', n)),
     xlab = expression(sigma["angular error"]), ylab = expression("E"["max"]))
legend("bottomleft", c("Random walk", "Directed walk", "Random rediscretized to length 1", "Random rediscretized to length 10"), pch = 16, col = c("red", "blue", "green", "black"), inset = 0.01)
```
